---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions,include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
```


## Loading and preprocessing the data

First we load the data and ensure the date fiels is properly parsed.
```{r}
stepData <- read.csv(unz("activity.zip","activity.csv"), header=TRUE, quote="\"")
stepData$date <- as.Date(stepData$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?

To calculate the average number of steps each day we must first find the total sum of steps for each day.
```{r}
dailyTotalSteps <- aggregate(steps ~ date, stepData, sum)
hist(dailyTotalSteps$steps,
     main="Daily Step Frequency",
     xlab="Number of Steps",
     col="red")
mean(dailyTotalSteps$steps)
median(dailyTotalSteps$steps)
```

The mean number of steps each day is `r formatC(mean(dailyTotalSteps$steps), format="d", big.mark=',')`, while
the median number of steps each day is `r formatC(median(dailyTotalSteps$steps), format="d", big.mark=',')`.


## What is the average daily activity pattern?

To analyze the pattern of daily activity, we must find the average number of steps for each 5-minute interval.
```{r}
intervalMeanSteps <- aggregate(steps ~ interval, stepData, mean)
plot(intervalMeanSteps$interval,
     intervalMeanSteps$steps,
     type="l",
     main="Average Steps by Time",
     xlab="Time (24-hour clock)",
     ylab="Average Number of Steps")
intervalMeanSteps[which.max(intervalMeanSteps$steps),1]
```

The 5-minute interval with the highest average number of steps
is `r intervalMeanSteps[which.max(intervalMeanSteps$steps),1]`.


## Imputing missing values

First calculate the number of rows in the dataset with missing values.
```{r}
sum(is.na(stepData$steps))
```

There are `r sum(is.na(stepData$steps))` rows in the dataset with missing Step values.

To impute a missing number of steps for a given interval, I will use the previously calculated
average number of steps for that interval across all dates, and use that value in place of the missing value.

```{r}
# Combine main data set with summarized interval data
stepDataImputed <- merge(stepData, intervalMeanSteps, by="interval")
colnames(stepDataImputed)[2] <- "steps"
colnames(stepDataImputed)[4] <- "intervalMeanSteps"

# Update NA values with mean value for the interval
for(i in 1:length(stepDataImputed$steps)) {
    if(is.na(stepDataImputed$steps[i])) {
        stepDataImputed$steps[i] <- stepDataImputed$intervalMeanSteps[i]
    }
}
```


Now we'll look at an updated histogram of daily steps taken

```{r}
dailyTotalStepsImputed <- aggregate(steps ~ date, stepDataImputed, sum)
hist(dailyTotalStepsImputed$steps,
     main="Daily Step Frequency",
     xlab="Number of Steps",
     col="red")
mean(dailyTotalStepsImputed$steps)
median(dailyTotalStepsImputed$steps)
```

The new mean number of steps each day is
`r formatC(mean(dailyTotalStepsImputed$steps), format="d", big.mark=',')`,
while the new median number of steps each day is
`r formatC(median(dailyTotalStepsImputed$steps), format="d", big.mark=',')`.
The mean and median were not noticably affected by this method of imputing missing values.

## Are there differences in activity patterns between weekdays and weekends?

First we must determine whether each row is a weekday or weekend.
```{r}
stepDataImputed$dayType <- weekdays(stepDataImputed$date)
stepDataImputed$dayType <- gsub("Saturday|Sunday","weekend",stepDataImputed$dayType)
stepDataImputed$dayType <- gsub("Monday|Tuesday|Wednesday|Thursday|Friday","weekday",stepDataImputed$dayType)
```

Now we summarize and plot the data.
```{r}
dayTypeSteps <- aggregate(stepDataImputed$steps,stepDataImputed[c(1,5)],mean)
colnames(dayTypeSteps)[3] <- "steps"

library(lattice)
xyplot(steps~interval | dayType,
       data = dayTypeSteps,
       type = 'l',
       xlab = 'Interval',
       ylab = 'Number of Steps',
       layout = c(1,2))
```
